---
title: "EMLeditor"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{EMLeditor}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```
# Dependencies

In addition to [EMLeditor](https://github.com/nationalparkservice/EMLeditor), you will also need the [EML](https://cran.r-project.org/web/packages/EML/index.html) package to complete these steps. 

You can download and install them individually, or get everything you need at once from the [NPSdataverse](https://github.com/nationalparkservice/NPSdataverse) using:

```{r verse_depends, eval=FALSE}
#install the NPSdataverse:
install.packages("devtools")
devtools::install_github("nationalparkservice/NPSdataverse")
library(NPSdataverse)
```

```{r ind_depends, eval=FALSE}
#individual install:
install.packages("devtools")
devtools::install_github("nationalparkservice/EMLeditor")
install.packages("EML")
library(EML)
library(EMLeditor)
```

# Workflow outline

EMLeditor’s primary objective is to edit and view EML formatted files, not to generate them from scratch. A suggested workflow is:

1) Use the EMLassemblyline::make_eml() to generate an initial EML document and save it as a .xml file (NPS naming convention is: *_metadata.xml)
2) Use the EML::read_eml() function to load your EML file into R as an R object.
3) Use EMLeditor functions to edit the metadata in R and evaluate whether your metadata is acceptable (don't forget to use EML::eml_validate() to make sure you are generating valid EML).
4) Use the EML::write_eml() function to write the R object back to XML (remember the NPS naming convention for metadata files is *_metadata.xml).

If you use EMLeditor functions to alter your metadata (e.g. any function with the prefix "set_" in the name) they will also silently add the National Park Service as a publisher (including location, [ROR id](https://ror.org/), etc) to your metadata unless you set NPS=FALSE. If you leave the default setting as NPS=TRUE, EMLeditor will also assume the data package is being created "by or for the NPS" and add that information to the metadata. 

"set_" functions will also inject information about which version of EMLeditor you used into the metadata.

# A Minimal Workflow

This workflow assumes your EML was generated using [EMLassemblyline](https://github.com/EDIorg/emlAssemblyLine), but will also work with any number of other EML generators (e.g. [ezEML](https://ezeml.edirepository.org/eml/)).

Currently, many of the steps in EMLeditor are interactive and will give you feedback on the current fields in your metadata, ask if you want to update them, and report on the results of your edits. 

### Import your metadata into R

```{r import, eval=FALSE}
my_metadata <- EML::read_eml("mymetadata_metdata.xml", from="xml")
```

### Add information about CUI

Add information about your Controlled Unclassified Information (CUI). Because it is important to indicate not only what type of CUI there is, but when (and why) there is not CUI, **you must do this even if your data package does not contain CUI**. Choose from one of seven CUI codes. These are:

  * PUBFUL - Does NOT contain CUI. The original data contained no CUI. No data were obscured or altered to generate the data package.
  * PUBVER - Does NOT contain CUI. The original data contained CUI, but in this data package CUI have been obscured so that it no longer contains CUI.
  * NPSONLY - Contains CUI. For NPS access only.
  * FED ONLY - Contains CUI. Only federal employees should have access (similar to "internal only" in DataStore).
  * FEDCON - Contains CUI. Only federal employees and federal contractors should have access (also very much like current "internal only" setting in DataStore)
  * DL ONLY - Contains CUI. Should only be available to a named list of individuals (where and how to list those individuals TBD)
  * NOCON - Contains  CUI. Federal, state, local, or tribal employees may have access, but contractors cannot.
  
The first three codes are NPS specific. More information about the remaining four codes can be found on the [National Archives website](https://www.archives.gov/cui/registry/limited-dissemination).

```{r cui, eval=FALSE}
my_meta2 <- set_cui(my_metadata, "PUBFUL")
```

### Add a data package DOI

Add your data package's Digital Object Identifier (DOI). To do this, you will need the seven-digit Reference ID from DataStore for the data package. To get the code, go to [DataStore](https://irma.nps.gov/DataStore/) and initiate a draft reference (but do not activate it). This act reserves your DOI, but it is not activated until after your data package is published.

There is no need to fill in any additional fields in DataStore at this point - many of them will automatically be populated based on your metadata at upload.

``` {r doi, eval=FALSE}
my_meta2 <- set_doi(my_meta2, 1234567)
```

### Add information about the DRR (optional)

If you are producing (or plan to produce) a DRR, add links to the DRR describing the data package. 

Similar to when you added the data package DOI, you will need the DOI for the DRR you are drafting as well as the DRR's Title. Again, go to [DataStore](https://irma.nps.gov/DataStore) and initiate a draft DRR, including a title. For the purposes of the data package, there is no need to populate any other fields. At this point, you do not need to activate the DRR reference and, while a DOI has been reserved for your DRR, it will not be activated until after publication so that you have plenty of time to construct the DRR.

```{r drr_doi, eval=FALSE}
my_meta2 <- set_drr_doi(my_meta2, 7654321, "DRR Title")
```

### Set the language

This is the human language (as opposed to computer language) that the data package and metadata are constructed in. Examples include English, Spanish, Navajo, etc. A [full list of available languages] is available from the Library of Congress. Please use the "English Name of Language" as an input. The function will then convert your input to the appropriate 3-character ISO 639-2 code. 

``` {r language, eval=FALSE}
my_meta2 <- set_language(my_meta2, "English")
```

### Add Content Unit Links

These are the park units where data were collected from, for instance ROMO, not ROMN. If the data package includes data from more than one park, they can all be listed. For instance, if data were collected from all park units within a network, each unit should be listed separately rather than the network. This is because the geographic coordinates corresponding to bounding boxes for each park unit listed will automatically be generated and inserted into the metadata. Individual park units will be more informative than the bounding box for the entire network.

``` {r park_units, eval=FALSE}
park_units <- c("ROMO", "GRSD", "YELL")
my_meta2 <- set_content_units(my_meta2, park_units)
```

### Add the Producing Unit

This is the unit responsible for generating the data package. It may be a single park (ROMO) or a network (ROMN). It may be identical to the units listed in the previous step, overlapping, or entirely different.

```{r prod_units, eval=FALSE}
my_meta2 <- set_producing_units(my_meta2, "ROMN")
```

Great! You're done adding the essential NPS-specific metadata to your EML. There are only two quick steps left:

### Validate your EML 

OK, this first one might take a tick to run:

```{r validate, eval=FALSE}
EML::eml_validate(my_meta2)
```

If eml_validate returns errors, inspect them and fix them. Feel free to contribute an [issue](https://github.com/nationalparkservice/EMLeditor/issues), or email [Rob Baker](mailto:robert_baker@nps.gov?subject=EMLeditor%20issues) with questions, concerns, or suggestions. 


### Write your edited EML back to disk

Assuming everything went smoothly and eml_validate returns 'TRUE', write your EML back to your disk so you can upload it with your data files to DataStore. Keep in mind that the file name should end in _metadata.xml. Also, when uploading your data package (data files and metadata) to DataStore, make sure to upload it to the correct draft reference!

```{r write, eval=FALSE}
EML::write_eml(my_meta2, "mymetadatafilename_metadata.xml")
```

# Scripting with EMLeditor

The interactive feedback and prompts provided by EMLeditor functions can be turned off to enable efficient scripting. All "set_" class functions have a parameter, _force_ that defaults to _force = FALSE_. To turn off the feedback and prompts, set _r force = TRUE_ when calling each function. Be careful using the functions in this way as they may - or may not - make changes to your metadata and you will not be advised of any change or lack of change. Inspect your final product carefully.

# non-NPS users

EMLeditor functions are designed primarily for use by staff at the National Park Service for publication of data packages to DataStore. Consequently, all "set_" class functions silently perform two operations by default:

1) They set the publisher to the National Park Service (and the location to the Fort Collins office) 
2) They specify the agency that created the data package as NPS and set a field "by or for NPS" to TRUE. 

You can prevent set_class functions from performing these operations by changing the default status of the parameter _NPS = TRUE_ to _NPS = FALSE_. This will leave your publisher information untouched and will not create an additionalMetadata item for the agency that created the data package.

If you would like to set the publisher to something other than the Fort Collins Office of the National Park Service or your would like to set the agency that created the data package to something other than NPS, use the _set_publisher_  function. Be sure to specify _NPS = FALSE_ or the function will perform the default operations (set publisher to NPS at the Fort Collins Office and set the agency to NPS).

**Warning:** set_publisher should only be used in a few, likely rare, circumstances:

1) If the publisher Is **NOT** the National Park Service 
2) If the contact address for the publisher is **NOT** the central office in Fort Collins (all data packages uploaded to DataStore will be published by the Fort Collins Office of NPS)
3) If the originating agency is **NOT** the NPS (i.e. a contractor or partner organization)
4) If the data package is **NOT** created for or by the NPS

It's probably a good idea to run
```{r args, eval=FALSE}
args(set_publisher)
```
to make sure you have all the arguments, especially those with defaults, properly specified.

# Check your EML

It's always a good idea to check your EML. Other than visually inspecting the .xml, file two good approaches are:

1) Check whether your EML is schema-valid:

```{r eml_validate, eval=FALSE}
# Use the eml_validate function from the EML package:
EML::eml_validate(mymetadata)

# Alternatively, use the eml validating tool in the DPchecker package.
# This performs the same checks with more intuitive feedback:
# First, write your EML to an xml file, following the file naming convention (ends with _metadata.xml)
EML::write_eml(mymetadata, "mymetadata_metadata.xml")

#Next, use DPchecker package to test whether your EML is schema-valid:
DPchecker::test_validate_schema()
```

2) Build a _mock-up_ of a readme file:

```{r readme, eval=FALSE}
# Outputs readme to the screen
write_readme(mymetadata) 

# Alternatively, save the readme to a text file
write_readme(mymetadata, "readme.txt") 
```

The _mock_ readme file is an approximation of the readme file that will automatically be generated by DataStore upon upload and included in your data package. It's a good, human readable, way to check whether many critical elements of your EML are properly formatted. Although the actual readme on DataStore may differ slightly, if the _mock_ readme looks good, that's a good indication that the readme on DataStore will too. On the other hand, if there is something off in the _mock_ readme, that's a good indication that you may want to go back and fix the relevant portions of your EML.

**do not upload your _mock_ readme file to DataStore!**
